Improved state-space models for inference about spatial and temporal variation in abundance from count data

### 2012-10-02 Dail_Mad_ext.tex
on to extensions

The DM model requires count data collected at $R$ sites, each of which
is surveyed on $T$ primary sampling periods. A site is ideally a
well-defined region of a study area, such as a wetland or a patch of
early-successional habitat, although it could be an arbitrarily
defined region such as a randomly located
survey plots. The timeframe of the study is arbitrary, but in general
it will be sufficiently long to allow for temporal variation in
abundance. For instance, a primary sampling period could be a one
month long breeding season, which could be surveyed once per year for
$T$ years. If the a site is surveyed on $J$ occassions during a
primary period, these are called secondary sampling occasions.
In the case where no secondary sampling was conducted, i.e. $J=1$,
let $n_{it}: i=1,\hdots,R; t=1,\hdots,T$ denote the count data at site
$i$ and primary period $t$. If secondary sampling periods were used,
an extra dimension is added so tha twe have $n_{ijt}:
j=1,\hdots,J$. Typically, we $n_{it}$ will be lower than the actual
quantitiy of interest, abundance,
denoted $N_{it}$. In cases where detection probability is perfect,
abundance is observed directly such that $n_{it} \equiv N_{it}$.

\subsection{The Model}

@@

Eqs.~\ref{eq:N1} and \ref{eq:Nt} fully specify the state model---the
model for spatial and temporal variation in abundance. The observation
model assumes that individuals are missed due to imperfect
detection. The simplest model for imperfect detection, and the one
considered in the original formulation, is
\begin{equation}
  n_{it} \sim \mathrm{Bin}(N_{it}, p)
  \label{eq:p1}
\end{equation}
Surprsingly detection probability, $p$, can be estimated without
secondary sampling occasions. However, this is only possible
if the parameteric assumptions of the population dynamics are met
exactly. In practice, it is preferable to obtain direct information
about $p$ using secondary sampling occassions or some other auxiliary
data, \hl{as is demonstrated in section....}

@@

Maximum likelihood estimation of parameters in models with random
effects requires removing the random effects by integration or
summation. For Markovian models, this can be done recursively, making
the likelihood tractible, the general expression of which can be written:
\begin{equation}
  \mathcal{L}(\Lambda, \Phi, \mathbf{p} | \{n_{it}\}) =
  \prod_{i=1}^R \left\{ \sum_{N_{i1}=0}^{\infty}
  [n_{i1}|N_{i1}][N_{i1}|\Lambda] \left\{
%  \prod_{t=T}^2 \left\{
  \sum_{N_{i1}=0}^{\infty} \dotsm \sum_{N_{iT}=0}^\infty
  [n_{it}|N_{it}][N_{it}|N_{it-1}] %\right\}
  \right\} \right\}
  \label{eq:like}
\end{equation}
This likelihood can be maximized numerically using the \textbf{R}
package \texttt{unmarked} \citep{fiske_chandler:2011}. Examples are
given in the Appendix.

@@

Writeing custom MCMC algorthims is often tedious and foreign to
ecologists, but software packages such as \textbf{WinBUGS} and
\textbf{JAGS} overcome the technical difficulties by allowing users to
specify the model using a simple symbolic descriptions. For example,
here is \textbf{BUGS} code for the ``constant'' model described above.

\begin{verbatim}
model {
lambda ~ dunif(0, 5)
omega ~ dunif(0,1)
gamma ~ dunif(0, 10)
p ~ dunif(0,1)
for(i in 1:nSites) {
  N[i,1] ~ dpois(lambda)
  y[i,1] ~ dbin(p, N[i,1])
  for(t in 2:nYears) {
    S[i,t-1] ~ dbin(omega, N[i,t-1])
    G[i,t-1] ~ dpois(gamma)
    N[i,t] <- S[i,t-1] + G[i,t-1]
    n[i,t] ~ dbin(p, N[i,1])
    }
  }
}
\end{verbatim}

>>>>

The DM model requires count data collected at $R$ sites, each of which
is surveyed on $T$ primary sampling periods. A site is ideally a
well-defined region of a study area, such as a wetland or a patch of
early-successional habitat, although it could be an arbitrarily
defined region such as a randomly located
survey plots. The timeframe of the study is arbitrary, but in general
it will be sufficiently long to allow for temporal variation in
abundance. For instance, a primary sampling period could be a one
month long breeding season, which could be surveyed once per year for
$T$ years. If the a site is surveyed on $J$ occassions during a
primary period, these are called secondary sampling occasions.
In the case where no secondary sampling was conducted, i.e. $J=1$,
let $X_{it}: i=1,\hdots,R; t=1,\hdots,T$ denote the count data at site
$i$ and primary period $t$. If secondary sampling periods were used,
an extra dimension is added so tha twe have $X_{ijt}:
j=1,\hdots,J$. Typically, we $X_{it}$ will be lower than the actual
quantitiy of interest, abundance,
denoted $X_{it}$. In cases where detection probability is perfect,
abundance is observed directly such that $X_{it} \equiv N_{it}$.

\subsection{The Original Formulation of the Model}

@@

Eqs.~\ref{eq:N1} and \ref{eq:Nt} fully specify the state model, i.e. the
model for abundance dynamics. This state
model has allowed us to replace the initial abundance distribution and
transition models of conventional state-space models, with
alternatives that (1) allow for modeling spatial variation in
abundance, (2) respect the discrete nature of the counts, and (3) have
parameters that are estimable. The one
remaining task is to replace the observation model
$[X_t|N_t, \bm{p}]$ (Eq.~\ref{eq:ss2c}) with one that can be more
easily interpretted. The observation
model assumes that individuals are missed due to imperfect
detection. The simplest model for imperfect detection, and the one
considered in the original formulation, is
\begin{equation}
  X_{it} \sim \mathrm{Bin}(N_{it}, p)
  \label{eq:p1}
\end{equation}
or equivalently,
\begin{equation}
  [X_{it} | N_{it}, \bm{p}] = \binom{N_{it}}{X_{it}} \bm{p}^{X_{it}}(1-\bm{p})^{N_{it}-X_{it}}.
  \label{eq:1}
\end{equation}
Under this model, $\bm{p}$ are interpretted as detection probability
parameters...

%%%% Save for discussion?
%Surprsingly detection probability, $p$, can be estimated without
%secondary sampling occasions. However, this is only possible
%if the parameteric assumptions of the population dynamics are met
%exactly. In practice, it is preferable to obtain direct information
%about $p$ using secondary sampling occassions or some other auxiliary
%data, \hl{as is demonstrated in section....}

The entire model can now be written:
\begin{gather}
  N_{i1} \sim \text{Pois}(\Lambda) \nonumber \\
  S_{it}|N_{it-1} \sim \mathrm{Bin}(N_{it-1}, \omega) \nonumber \\
  G_{it}|N_{it-1}\sim \mathrm{Pois}(\gamma(N_{it-1})) \\
  N_{it} = S_{it}+G_{it} \nonumber \\
  X_{it} \sim \text{Bin}(N_{it}, p) \nonumber
  \label{eq:hm}
\end{gather}

@@

Maximum likelihood estimation of parameters in models with discrete random
effects requires removing them by
summation. For Markovian models, this can be done recursively, making
the likelihood tractible. A general expression of the likelihood can be written:
\begin{multline}
  \mathcal{L}(\Lambda, \bm{\Theta}, p | \{X_{it}\}) = \\
  \prod_{i=1}^R \left\{ \sum_{N_{i1}=0}^{\infty}
  [X_{i1}|N_{i1}, p][N_{i1}|\Lambda] \left\{
%  \prod_{t=T}^2 \left\{
  \sum_{N_{i2}=0}^{\infty} \dotsm \sum_{N_{iT}=0}^\infty
  [X_{it}|N_{it}, p][N_{it}|N_{it-1}, \bm{\Theta}] %\right\}
  \right\} \right\}
  \label{eq:like}
\end{multline}
where $\infty$ is replaced by a finite upper bound $K$, which should
be high enough such that the MLEs are not affected by increasint it
further. This likelihood can be maximized numerically using the \textbf{R}
package \texttt{unmarked} \citep{fiske_chandler:2011}. Examples are
given in the Appendix.

@@

Writeing custom MCMC algorthims is often tedious and foreign to
ecologists, but software packages such as \textbf{WinBUGS} and
\textbf{JAGS} overcome the technical difficulties by allowing users to
specify the model using a simple symbolic descriptions. For example,
here is \textbf{BUGS} code for the ``geometric'' model described shown
in Eq.~\ref{eq:hm}:
\begin{verbatim}
model {
lambda ~ dunif(0, 5)
omega ~ dunif(0,1)
gamma ~ dunif(0, 10)
p ~ dunif(0,1)
for(i in 1:nSites) {
  N[i,1] ~ dpois(lambda)
  y[i,1] ~ dbin(p, N[i,1])
  for(t in 2:nYears) {
    S[i,t-1] ~ dbin(omega, N[i,t-1])
    G[i,t-1] ~ dpois(gamma*N[i,t-1])
    N[i,t] <- S[i,t-1] + G[i,t-1]
    n[i,t] ~ dbin(p, N[i,1])
    }
  }
}
\end{verbatim}
